Seminar: Programming models and Code Generation\\ Winter Term 2014/2015\\ {\bf Quasi deterministic parallel programming using LVars}

### 2014-12-26 draft.tex
bibliography

There is a need to generalize the single-assignment model to something that admits programs like these but retains determinism. Deterministic algorithms must be expressible in a style that guarantees determinism, and non-deterministic behaviors, where desired, must be requested explicitly. More generally, enforcing deterministic semantics simplifies composing, porting, reasoning about, debugging, and testing parallel software. Language-level enforcement of determinism is possible, but existing models tend to lack features that would make them applicable to a broader range of problems. Moreover, they lack extensibility. \\ \\

@@

% free floating figure using width of full page, to be put on [t]op.
\begin{figure}
\centerline{
\includegraphics[width=1.0\columnwidth]{Figures/lattice.jpg}
}
\caption{A nice caption. The larger width allows for more text without
taking too much space.}
\label{Fig2}
\end{figure}
\begin{itemize}
\item $D$ has a least element $\bot$, representing the initial “empty” state of an Lvar,
\item $D$ has a greatest element $\top$, representing the “error” state that results from conflicting updates to an Lvar,
\item $D$ comes equipped with a partial order $\userleq$, where $\bot \userleq d \userleq \top$ for all $d \in D$.
\item Every pair of elements in $D$ has a least upper bound (lub) written $\sqcup$ which means it is possible for two sub-computations to independently update an LVar, and then deterministically merge the results by taking the lub of the resulting two states.
\end{itemize}
Data structures like arrays, tree, maps , streams etc. can be represented  as lattice. $Figure$ $2$ gives three examples of lattices for common data structures.

@@

For Lambda Lvar syntax and semantics, remaining definitions and its details please refer XXX. 

@@

\textbf{\textit{Proof of determinism:}} The key to the determinism is the frame rule given in \emph{Figure 2}. A frame property captures the idea of local reasoning about programs that alter state. Here, C is a program, and $\{p\} C \{q\}$ says that if the assertion p  is true before C  runs, then the assertion q  will be true afterwards. For example, $p$ and $q$  might describe the state of the heap before and after it is updated by $C$. It tells that running $C$ starting from a state satisfying the assertion $p$ $∗$ $r$ will result in a state satisfying the assertion $q$ $∗$ $r$. The assertion $p$ $*$ $r$ is satisfied by a heap if the heap can be split into two non-overlapping parts satisfying $p$ and $r$, respectively.

@@

\textbf{\textit{Diamond Lemma:}} This does the heavy lifting of the determinism proof: it establishes the diamond property, which says that if a configuration steps to two different configurations, there exists a single third configuration to which those configurations both step. For supporting lemmas, its details and complete proofs please refer to XXXXXXXXXX. Lemmas there also show  how cases of deadlocks and livelocks are also handled.

@@

In this regard, Lvish extends the Lvars model with 2 additional capabilities. First, it adds event handlers, a mechanism for attaching a callback function to an LVar that runs, asynchronously, whenever events arrive (in the form of monotonic updates to the Lvar). Second, a primitive for freezing an LVar, which means: once an LVar is frozen, any further writes that would change its value instead throw an exception; on the other hand, it becomes possible to discover the exact value of the LVar, learning both positive and negative information about it, without blocking.

@@

[Paper] contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a store containing LVars. It extends the original LVar formalism to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern—which is a simpler primitive combining all others.

An LVar’s state is now a pair $(d,frz )$, where $d$ is an element of the application-specific set $D$ and $frz$(Short for freeze) is a “status bit” of either $true$ or $false$. We define an ordering $\leqp$ on LVar states $\state{d}{\status}$ in terms of the application-specific ordering $\userleq$ on elements of $D$. Every element of $D$ is ``freezable'' except $\top$. Informally: 
\begin{itemize}
\item Two unfrozen states are ordered according to the application-specific $\userleq$; that is, $(d,false) \leqp (d',false)$
exactly when $d \userleq d'$.
\item Two frozen states do not have an order, unless they are equal: $(d,true) \leqp (d',true)$ exactly when $d = d'$.
\item An unfrozen state $(d,false)$ is less than or equal to a frozen state $(d',true)$ exactly when $d \userleq d'$.
\item A frozen state is less than an unfrozen state if the unfrozen state is $\top$; that is, $(d,true) \leqp (d',false)$ exactly when $d' = \top$.
\end{itemize}
The addition of status bits to the application-specific lattice results in a new lattice $(D_p, \leqp, \botp, \topp)$. $\lubp{}{}$ for the lub operation that $\leqp$ induces. During the evaluation of LVish programs, a \emph{store} $S$ keeps track of the states of LVars and represents them by a binding from a location $l$, drawn from a set $loc$, to its state, which is some pair $(d, frz)$ from the set $D_p$. For detailed syntax and semantics of Lvish please refer to the paper[7].\\ \\
\textbf{\textit{Communication primitives:}} They are same as Lvar except that new will now create a binding for a new Lvar whose initial state is (?, false). Put takes a pointer to the Lvar and a new lattice element which is a pair. Get similarly takes a threshold set containing pairs. $freeze\ e_{lv}\ after\ e_{events}\ with\ e_{cb}$ has the following semantics: It attaches the callback $e_{cb}$ to the LVar $e_{lv}$. The expression $e_{events}$ must evaluate to a event set $Q$; the callback will be executed, once, for each lattice element in $Q$ that the LVar’s state reaches or surpasses. The callback takes a lattice element as its argument. Its return value is ignored, so it runs solely for effect. For instance, a callback might itself do a put to the LVar to which it is attached, triggering yet more callbacks. If the handler reaches a quiescent state, the LVar $e_{lv}$ is frozen, and its exact state is returned (rather than an under-approximation of the state, as with $get$).

To keep track of the running callbacks, LVish includes an auxiliary form of the above pattern taking additional arguments: the set of running callbacks $B$ and a set of values $H$ in the event set for which callbacks have already been launched. It detects quiescence by checking: Every event of interest that has occurred must be handled (be in $H$). Second, all existing callback threads must have completed and terminated with a value. If yes, then LVars state is freezed. To ensure quiescence is a permanent state, rather than a transient one, freezing is usually the very last step of an algorithm, permitting its result to be extracted. The $runParThenFreeze$ function inside LVish library does this, and thereby guarantees full determinism.\\ \\
\textbf{\textit{Proof of Quasi determinism for Lvish:}} According to the paper[7] the main theorem says that if two executions starting from a configuration $\sigma$ terminate in configurations $\sigma{'}$ and $\sigma{''}$, then either $\sigma{'}$ and $\sigma{''}$ are the same configuration (up to a permutation on locations), or one of them is \emph{error}.\\
\textit{Independence property:} Like LVars, the LVish model uses the frame property again to capture the idea that independent effects commute with each other but this time with some restrictions. If we know that a \emph{store} $S$ evaluating expression $e$ steps to a new \emph{store} $S{'}$ with $e{'}$ then it is valid to say that $S\sqcup_{S}S''$ evaluating $e$ steps to a new store $S'\sqcup_{S}S''$ with $e'$ if the following conditions are satisfied:\\
\begin{itemize}
\item The status bits of $S'\sqcup_{S}S''$ and S agree with each other for all locations shared between them.
\item Locations in $S''$ cannot share the same name with locations newly allocated during the transition from one configuration
to another.
\end{itemize}
Please refer to the original paper and the technical report for elaborate details on lemmas and their proofs.

@@

It adopts the basic approach of the par monad. The existing Par monad[3] in Haskell offers an API for parallel programming. It provides by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With

@@

 In order to specify the type of the par computation it provides a phantom type that indicates the determinism level.\\ \\

@@

following suite of run functions\\ \\

@@

runPar : : Par Det a -> a\\

@@

runParThenFreeze : : Deep frz a =>\\

@@

\textit{Idempotence,} meaning that $d\sqcup{d=d}$ for any element $d$. Repeated puts or freezes have no effect, if the scheduler is allowed to occasionally duplicate work, it is possible to save on synchronization costs. Since LVish computations are guaranteed to be idempotent, we could use such a scheduler (standard Chase-Lev deque [10]). Idempotence helps to deal with races between $put$ and $get/addHandler$. \\ \\

@@

The bag module supports atomic insertion and removal, and concurrent traversal.\\

@@

A listener for an LVar is a pair of callbacks, one called when the LVars lattice value changes, and the other when the LVar is frozen. \\

@@

Internally, the Par monad represents computations in continuation-passing style, in terms of their interpretation in the IO monad.\\

@@

\textbf{\textit{Evaluation: k-CFA study}} In [7] a case study on k-CFA (parallelizing control flow analysis) is done to evaluate the performance of the Lvish implementation. K-CFA provide a hierarchy of increasingly precise methods to compute the control 

@@

\textbf{\textit{Related work:}} In [6], feasibility of LVar threshold reads is explored in a distributed setting. Distributed systems involve replication of data objects across several locations making them robust to data loss and give better data locality. Eventual consistency is a policy in which the updates eventually reach the replicas and the replicas need not agree at all times. Convergent replicated data types provide a simple mathematical framework for reasoning about and enforcing the eventual consistency of such replicated objects, based on viewing them as elements of a lattice and replica conflict resolution as the lattice's join operation. CvRDTs are provably eventually consistent, but queries of CvRDTs allow inconsistent intermediate

@@

% free floating figure using width of full page, to be put on [t]op.
\begin{figure}
\centerline{
\includegraphics[width=1.3\columnwidth]{Figures/speedup.jpg}
}
\caption{A nice caption. The larger width allows for more text without
taking too much space.}
\label{Fig2}
\end{figure}

\section{Summary}
\label{summary}
In an attempt towards creating better deterministic parallel programming models, Lvish has extended the existing usage of monotonicity as a theme to provide better applicability, usability, performance, providing a single generic model that subsumes
many of the single assignment models, thereby representing a larger class of problems which can be effciently parallelized. The performance with regards to certain benchmark applications are an evidence of the same. [4]discusses the viability of Lvish for certain graph algorithms, its drawbacks and proposed extension. Some of the drawbacks to highlight are the fact that Lvish side  effects are only monotonic updates and if a data structure needs to be updated multiple times it will have to be done out of
place thereby frequently allocating LVars becoming a source of additional memory allocation and ineffcient relative to in-place versions. Blocking $get$ semantics could slow down until the dependency is met reducing the performance, hence instead
of blocking, the tasks could be retried at a later time/rescheduled when there dependency has been satisfied. The overhead in going through the scheduler many times has also been highlighted in[4] for applications like Maximum independent set. The
paper proposes a solution similar to an abort/retry feature.
% Put citations from bibtex into References section which were not
% explicity cited.
\nocite{robotron,
stonx,vice,650sim,herculessim,zib,4004,thermal1,thermal2,rojas,neumann,
neumann1void,neumann2void,neumann3,neumann4,Siewiorek,AmBlBr,Blaauw:1997:CAC,ChPB97,Brom98,Clym93,Buhu99,
Edwa01,Nill99,ABC+90,Rama91,Heid97,Kist95,Klot03,LMcCL92,LFS+92,LMDD92,Vlec03,Cray77,top500,650,Amda67,
Arla88,ODLD86,Hick04,Walk04}

\bibliographystyle{plain}
% Literature sources are to be found in seminarpaper.bib
\bibliography{seminarpaper} 
\end{document}

>>>>

There is a need to generalize the single-assignment model to admits programs like these but retains determinism. Deterministic algorithms must be expressible in a style that guarantees determinism, and non-deterministic behaviors, where desired, must be requested explicitly. Enforcing deterministic semantics simplifies composing, porting, reasoning about, debugging, and testing parallel software. Existing models tend to lack features that would make them applicable to a broader range of problems. Moreover, they lack extensibility. \\ \\

@@

% free floating figure using width of full page, to be put on [t]op.
\begin{figure}
\centerline{
\includegraphics[width=1.0\columnwidth]{Figures/lattice.jpg}
}
\caption{Example lattices\cite{lkuper}: (a) Positive integers ordered by $\leq$ (b) IVar containing a natural number (c) Pair of natural number-valued IVars}
\label{Figure1}
\end{figure}
\begin{itemize}
\item $D$ has a least element $\bot$, representing the initial “empty” state of an Lvar,
\item $D$ has a greatest element $\top$, representing the “error” state that results from conflicting updates to an Lvar,
\item $D$ comes equipped with a partial order $\userleq$, where $\bot \userleq d \userleq \top$ for all $d \in D$.
\item Every pair of elements in $D$ has a least upper bound (lub) written $\sqcup$ which means it is possible for two sub-computations to independently update an LVar, and then deterministically merge the results by taking the lub of the resulting two states.
\end{itemize}
Data structures like arrays, tree, maps , streams etc. can be represented  as lattice. $Figure$ $2$ gives three examples of lattices for common data structures.

@@

For Lambda Lvar syntax and semantics, remaining definitions and its details please refer\cite{lkuper}. 

@@

\textbf{\textit{Proof of determinism:}} The key to the determinism is the frame rule given in \cite{Figure1}. A frame property captures the idea of local reasoning about programs that alter state\cite{lindsey}. Here, C is a program, and $\{p\} C \{q\}$ says that if the assertion p  is true before C  runs, then the assertion q  will be true afterwards. For example, $p$ and $q$  might describe the state of the heap before and after it is updated by $C$. It tells that running $C$ starting from a state satisfying the assertion $p$ $∗$ $r$ will result in a state satisfying the assertion $q$ $∗$ $r$. The assertion $p$ $*$ $r$ is satisfied by a heap if the heap can be split into two non-overlapping parts satisfying $p$ and $r$, respectively.

@@

\textbf{\textit{Diamond Lemma:}} This does the heavy lifting of the determinism proof: it establishes the diamond property, which says that if a configuration steps to two different configurations, there exists a single third configuration to which those configurations both step. For supporting lemmas, its details and complete proofs please refer to\cite{lkuper}. Lemmas there also show how cases of deadlocks and livelocks are also handled.

@@

In this regard, Lvish\cite{lkuper2} extends the Lvars model with 2 additional capabilities. First, it adds event handlers, a mechanism for attaching a callback function to an LVar that runs, asynchronously, whenever events arrive (in the form of monotonic updates to the Lvar). Second, a primitive for freezing an LVar, which means: once an LVar is frozen, any further writes that would change its value instead throw an exception; on the other hand, it becomes possible to discover the exact value of the LVar, learning both positive and negative information about it, without blocking.

@@

\cite{lkuper2}contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a store containing LVars. It extends the original LVar formalism to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern—which is a simpler primitive combining all others.

An LVar’s state is now a pair $(d,frz )$, where $d$ is an element of the application-specific set $D$ and $frz$(Short for freeze) is a “status bit” of either $true$ or $false$. We define an ordering $\leqp$ on LVar states $\state{d}{\status}$ in terms of the application-specific ordering $\userleq$ on elements of $D$. Every element of $D$ is ``freezable'' except $\top$. Informally: 
\begin{itemize}
\item Two unfrozen states are ordered according to the application-specific $\userleq$; that is, $(d,false) \leqp (d',false)$
exactly when $d \userleq d'$.
\item Two frozen states do not have an order, unless they are equal: $(d,true) \leqp (d',true)$ exactly when $d = d'$.
\item An unfrozen state $(d,false)$ is less than or equal to a frozen state $(d',true)$ exactly when $d \userleq d'$.
\item A frozen state is less than an unfrozen state if the unfrozen state is $\top$; that is, $(d,true) \leqp (d',false)$ exactly when $d' = \top$.
\end{itemize}
The addition of status bits to the application-specific lattice results in a new lattice $(D_p, \leqp, \botp, \topp)$. $\lubp{}{}$ for the lub operation that $\leqp$ induces. During the evaluation of LVish programs, a \emph{store} $S$ keeps track of the states of LVars and represents them by a binding from a location $l$, drawn from a set $loc$, to its state, which is some pair $(d, frz)$ from the set $D_p$. For detailed syntax and semantics of Lvish please refer to the paper\cite{lkuper2}.\\ \\
\textbf{\textit{Communication primitives:}} They are same as Lvar except that new will now create a binding for a new Lvar whose initial state is (?, false). Put takes a pointer to the Lvar and a new lattice element which is a pair. Get similarly takes a threshold set containing pairs. $freeze\ e_{lv}\ after\ e_{events}\ with\ e_{cb}$ has the following semantics: It attaches the callback $e_{cb}$ to the LVar $e_{lv}$. The expression $e_{events}$ must evaluate to a event set $Q$; the callback will be executed, once, for each lattice element in $Q$ that the LVar’s state reaches or surpasses. The callback takes a lattice element as its argument. Its return value is ignored, so it runs solely for effect. For instance, a callback might itself do a put to the LVar to which it is attached, triggering yet more callbacks. If the handler reaches a quiescent state, the LVar $e_{lv}$ is frozen, and its exact state is returned (rather than an under-approximation of the state, as with $get$).

To keep track of the running callbacks, LVish includes an auxiliary form of the above pattern taking additional arguments: the set of running callbacks $B$ and a set of values $H$ in the event set for which callbacks have already been launched. It detects quiescence by checking: Every event of interest that has occurred must be handled (be in $H$). Second, all existing callback threads must have completed and terminated with a value. If yes, then LVars state is freezed. To ensure quiescence is a permanent state, rather than a transient one, freezing is usually the very last step of an algorithm, permitting its result to be extracted. The $runParThenFreeze$ function inside LVish library does this, and thereby guarantees full determinism.\\ \\
\textbf{\textit{Proof of Quasi determinism for Lvish:}} According to the paper\cite{lkuper2} the main theorem says that if two executions starting from a configuration $\sigma$ terminate in configurations $\sigma{'}$ and $\sigma{''}$, then either $\sigma{'}$ and $\sigma{''}$ are the same configuration (up to a permutation on locations), or one of them is \emph{error}.\\
\textit{Independence property:} Like LVars, the LVish model uses the frame property again to capture the idea that independent effects commute with each other but this time with some restrictions. If we know that a \emph{store} $S$ evaluating expression $e$ steps to a new \emph{store} $S{'}$ with $e{'}$ then it is valid to say that $S\sqcup_{S}S''$ evaluating $e$ steps to a new store $S'\sqcup_{S}S''$ with $e'$ if the following conditions are satisfied:
\begin{itemize}
\item The status bits of $S'\sqcup_{S}S''$ and S agree with each other for all locations shared between them.
\item Locations in $S''$ cannot share the same name with locations newly allocated during the transition from one configuration
to another.
\end{itemize}
Please refer to the original paper and the technical report for elaborate details on lemmas and their proofs.

@@

It adopts the basic approach of the par monad. The existing Par monad\cite{simon} in Haskell offers an API for parallel programming. It provides by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With

@@

In order to specify the type of the par computation it provides a phantom type that indicates the determinism level.

@@

following suite of run functions

@@

runPar : : Par Det a -> a

@@

runParThenFreeze : : Deep frz a =>

@@

\textit{Idempotence,} meaning that $d\sqcup{d=d}$ for any element $d$. Repeated puts or freezes have no effect, if the scheduler is allowed to occasionally duplicate work, it is possible to save on synchronization costs. Since LVish computations are guaranteed to be idempotent, we could use such a scheduler (standard Chase-Lev deque\cite{sched}). Idempotence helps to deal with races between $put$ and $get/addHandler$. \\ \\

@@

The bag module supports atomic insertion and removal, and concurrent traversal.

@@

A listener for an LVar is a pair of callbacks, one called when the LVars lattice value changes, and the other when the LVar is frozen. 

@@

Internally, the Par monad represents computations in continuation-passing style, in terms of their interpretation in the IO monad.

@@

\textbf{\textit{Evaluation: k-CFA study}} In \cite{lkuper2} a case study on k-CFA (parallelizing control flow analysis) is done to evaluate the performance of the Lvish implementation. K-CFA provide a hierarchy of increasingly precise methods to compute the control 

@@

\textbf{\textit{Related work:}} In \cite{newton2}, feasibility of LVar threshold reads is explored in a distributed setting. Distributed systems involve replication of data objects across several locations making them robust to data loss and give better data locality. Eventual consistency is a policy in which the updates eventually reach the replicas and the replicas need not agree at all times. Convergent replicated data types provide a simple mathematical framework for reasoning about and enforcing the eventual consistency of such replicated objects, based on viewing them as elements of a lattice and replica conflict resolution as the lattice's join operation. CvRDTs are provably eventually consistent, but queries of CvRDTs allow inconsistent intermediate

@@

% free floating figure using width of full page, to be put on [t]op.
\begin{figure}
\centerline{
\includegraphics[width=1.3\columnwidth]{Figures/speedup.jpg}
}
\caption{Performance of LVish on parallelizing k-CFA\cite{lkuper2}}
\label{Figure2}
\end{figure}

\section{Summary}
\label{summary}
In an attempt towards creating better deterministic parallel programming models, Lvish has extended the existing usage of monotonicity as a theme to provide better applicability, usability, performance, providing a single generic model that subsumes
many of the single assignment models, thereby representing a larger class of problems which can be effciently parallelized. The performance with regards to certain benchmark applications are an evidence of the same. \cite{newton1}discusses the viability of Lvish for certain graph algorithms, its drawbacks and proposed extension. Some of the drawbacks to highlight are the fact that Lvish side  effects are only monotonic updates and if a data structure needs to be updated multiple times it will have to be done out of
place thereby frequently allocating LVars becoming a source of additional memory allocation and ineffcient relative to in-place versions. Blocking $get$ semantics could slow down until the dependency is met reducing the performance, hence instead
of blocking, the tasks could be retried at a later time/rescheduled when there dependency has been satisfied. The overhead in going through the scheduler many times has also been highlighted in[4] for applications like Maximum independent set. The
paper proposes a solution similar to an abort/retry feature.
% Put citations from bibtex into References section which were not
% explicity cited.
\bibliographystyle{plain}
% Literature sources are to be found in seminarpaper.bib
\bibliography{seminarpaper} 
\end{document}
